---
title: "Factor Analysis"
author: "Huey Kwik & Richard Zhang"
date: "May 17, 2016"
output: pdf_document
---

Solutions adapted from work by Huey Kwik & Richard Zhang (Signal Cohort #2).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('dplyr')
library('psych')
library('GPArotation')
library('corrplot')

set.seed(1)
```

## Factor Analysis

Let's create a factors data frame!
```{r factors data frame}
X = rnorm(100)
Y = rnorm(100)
Z = rnorm(100)

factors = data.frame(X,Y,Z)
colnames(factors) = c("X","Y","Z")

noisyIndicators = function(feature, k, correlation) {
  noisies = lapply(1:k, function(x) {
    error = rnorm(length(feature))
    d = sqrt(1-correlation^2)
    correlation * feature + d * error
  })
  return(noisies)
}

xProxies = noisyIndicators(X, 4, 0.9)
yProxies = noisyIndicators(Y, 3, 0.9)
noisies = data.frame(xProxies, yProxies)
colnames(noisies) = as.character(1:7)
```

Check out the correlations!

```{r}
corrplot(cor(noisies), is.corr = TRUE)
cor(cbind(X, noisies))
```
 Run PCA on the noisy dataframe

```{r PCA}

pca = prcomp(noisies, scale=TRUE)
cor_pca = cor(cbind(pca$x,noisies))
corrplot(cor_pca, is.corr=TRUE)
```

##Part 2: Orthogonal Factor Analysis

```{r factor analysis}
pca = prcomp(noisies, scale=TRUE)
cor_pca = cor(cbind(pca$x,noisies))
corrplot(cor_pca, is.corr=TRUE)
```


Generate Variables
```{r generates 50 variables of length 100, warning=FALSE}
vars = sapply(1:50, function(i) { X*runif(1)+Y*runif(2)+Z*runif(3)+0.5*rnorm(1) })
pca = prcomp(vars, scale=TRUE)
```
Principal component 1 picks up on some of X Y Z
Principal component 2 picks up on more of X than Y and Z
Principal component 3 picks up on Y and Z

Corrplot of Principal Components with X,Y,Z
```{r corrplot of pca of variables with X Y Z}
cor_pca = cor(cbind(pca$x[,1:3], X,Y,Z))
corrplot(cor_pca, is.corr=TRUE, type="upper") 
```

MR1 mostly picks up on Y
MR2 mostly picks up on X
MR3 mostly picks up on Z
```{r corrplot of factors, warning=FALSE, message=FALSE}

fac_f = fa(vars, nfactors=3, rotate="varimax")
cor_fac = cor(cbind(fac_f$scores, X,Y,Z))
corrplot(cor_fac, is.corr=TRUE, type="upper") 
```


##Part 3: Oblique Factor Analysis

Create a noisy dataframe!
```{r noisy dataframe with X and W}

W = 0.5*X + Y

cor(W,Y) 

xProxies = noisyIndicators(X, 10, 0.8)
wProxies = noisyIndicators(W, 4, 0.8)
noisies = data.frame(xProxies, wProxies)
colnames(noisies) = as.character(1:14)

corrplot(cor(noisies), is.corr = TRUE)
```

Compare varimax rotation with oblimin rotation
```{r corrplot comparisons of rotation methods}
corrplot(cor(cbind(X, W, noisies)), is.corr=TRUE, type="upper")

library('GPArotation')
fac_f = fa(noisies, nfactors=2, rotate="varimax")
fac_obli = fa(noisies, nfactors=2, rotate="oblimin")

cor_fac_noisies = cor(cbind(fac_f$scores, noisies))
corrplot(cor_fac_noisies, is.corr=TRUE, type="upper") 

cor_fac = cor(cbind(fac_f$scores, X,W,Y))
corrplot(cor_fac, is.corr=TRUE, type="upper") 

cor_fac_obli = cor(cbind(fac_obli$scores, X,W,Y))
corrplot(cor_fac_obli, is.corr=TRUE, type="upper") 
```
 
Compared to varimax:
 
* MR1 is more highly correlated with MR2 
* MR1 and X are more highly correlated
* MR2 and W are more highly correlated 
 
## Speed Dating Data
 
Our goal is to detect factors in the speed dating dataset.

```{r}
df = read.csv("C:/Users/Andrew/Documents/Signal/curriculum/datasets/speed-dating/speeddating-aggregated.csv")
activities = select(df, sports:yoga)
fac_activities = fa(activities, nfactor=1, rotate='varimax')
corrplot(fac_activities$loadings, is.corr=FALSE)
```

Corrplots with 1-4 factors, comparing varimax vs oblique rotation.
 
```{r}
fac = function(n, rotation){
  fac_activities = fa(activities, nfactor=n, rotate=rotation)
  corrplot(fac_activities$loadings, is.corr=FALSE)
}
fac(2,"varimax")
fac(3,"varimax")
fac(4,"varimax")
fac(1,"oblimin")
fac(2,"oblimin")
fac(3,"oblimin")
fac(4,"oblimin") # Splits nicely into four factors!
```

## Big Five Personality Data

Compare Principal components and 5 factors analysis
```{r}
df = read.csv("C:/Users/Andrew/Documents/Signal/curriculum/datasets/big-5/data.csv", sep="\t")
questions = select(df, E1:O10)
pca_5 = prcomp(scale(questions))
fac_5 = fa(questions, nfactor=5, rotate="varimax")

corrplot(pca_5$rotation[,1:5], is.corr=FALSE)


corrplot(fac_5$loadings, is.corr=FALSE)
```

The former is very noisy, while the latter cleanly indicates 5 factors.

Regression Analysis: Predicting Gender using Questions

```{r}
gender_questions = cbind(df$gender, questions)
colnames(gender_questions)[1] = "gender"
gender_questions = dplyr::filter(gender_questions, gender != 3)
gender_questions$gender[gender_questions$gender == 2] = 0

fit = glm(gender ~ ., data = gender_questions, family = "binomial")
summary(fit) # Many many questions seem to be correlated with gender 
```
  
Each question predicts gender poorly (low coefficients)

Factor Analysis: Predicting Gender using 5 Factors
```{r}
gender_factors = as.data.frame(cbind(df$gender, fac_5$scores))
colnames(gender_factors)[1] = "gender"
gender_factors = dplyr::filter(gender_factors, gender != 3)
gender_factors$gender[gender_factors$gender == 2] = 0


fit = glm(gender ~ ., data = as.data.frame(gender_factors), family = "binomial")
summary(fit) 
```
Each factor predicts gender with a comparatively large coefficient.

Identifying the 5 factors from the questions:

* E1-E10 -> Extraversion
* N1-N10 -> Neuroticism
* A1-A10 -> Agreeableness
* C1-C10 -> Conscientiousness
* O1-O10 -> Openness

## Identifying Subfactors
```{r}
subfactors = function(questions) {
  fac_2 = fa(questions, nfactor=2, rotate="varimax")
  fac_3 = fa(questions, nfactor=3, rotate="varimax")
  fac_4 = fa(questions, nfactor=4, rotate="varimax")
  fac_5 = fa(questions, nfactor=5, rotate="varimax")
  
  # corrplot
  corrplot(fac_2$loadings, is.corr=FALSE) 
  corrplot(fac_3$loadings, is.corr=FALSE) 
  corrplot(fac_4$loadings, is.corr=FALSE)
  corrplot(fac_5$loadings, is.corr=FALSE) 
}

subfactors(select(df, E1:E10))
```

Extraversion:

* 2 factor - Separates out E8/E9, but not coherent
* 3 factor - MR1 - talking to people, MR2 - attention, MR3 - talking
* 4 factor MR1 - life of the party, MR2 - talking to people, MR3 - attention, MR4 - talking
* 5 factor - Additional factor doesn't seem to add much value

```{r}
subfactors(select(df, N1:N10))
```

Neuroticism: 

* 2 factor - MR1 - mood swings, MR2 - not coherent
* 3 factor - MR1 - mood swings, MR2 - irritability, MR3 - not coherent
* 4 factor - MR1 - mood swings, MR2 - irritability, MR3 - depression, MR4 - irritability
* 5 factor - MR1 - mood swings?, MR2 - mood swings?, MR3 - worry/stress, MR4 - depression, MR5 - upset/irritated

```{r}
subfactors(select(df, A1:A10))
```

Agreeability:

* 2 factor - MR1 - interest in others, MR2 - empathy
* 3 factor - MR1 - empathy, MR2 - interest in others, MR3 - interest in tohers?
* 4 factor - MR1- empathy, MR2 - interest in others, MR3 - make people feel at ease, MR4 - insult people
* 5 factor - MR1 - interest in people, MR2 - make people feel at ease, MR3 - empathy, MR4 - interest in people, MR5 - insult people

```{r}
subfactors(select(df, C1:C10))
```

Conscientiousness:

* 2 factor - MR1 - responsibility?, MR2 - tidy/messy
* 3 factor - MR1 - tidy/messy, MR2 - responsibility, MR3 - lack of responsiblity
* 4 factor - MR1 - schedule, MR2 - messy, MR3 - detail-oriented, MR4 - lack of responsibilty
* 5 factor - MR1 - messy, MR2 - forget messy, MR3 - schedule, MR4 - detail oriented, MR5 - lack of responsibility

```{r}
subfactors(select(df, O1:O10))
```

Openness:

* 2 factor - MR1 - imagination, MR2 - vocabulary
* 3 factor - MR1 - ideas, MR2 - vocabulary, MR3 - abstract ideas
* 4 factor - MR1 - abstract ideas, MR2 - good ideas?, MR3 - vocabulary, MR4 - imagination 
* 5 factor - MR1 - ideas, MR2 - vocabulary, MR3 - abstract ideas, MR4 - imagination, MR5 - learning?
